---
title: "Cookbook"
author: "Adam H. Sparks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{dplyr}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{magick}
  %\VignetteDepends{gapminder}
  %\VignetteDepends{scales}
  %\VignetteDepends{tibble}
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5
)
```

## How to Create AAGI Style Graphics

The CCDM CBADA team has developed an R package, {theme.aagi}, and this R cookbook to help ease the process of creating publication-ready graphics in our in-house style using both R's {graphics} and {ggplot2} libraries a more reproducible process, as well as making it easier for people new to R to create beautiful graphics that adhere to AAGI style guidelines.

### On {ggplot2} Plots and Graphs

Most of the focus of {theme.aagi} is given to supporting {[ggplot2](https://ggplot2.tidyverse.org/R)} visualisations, but {base} and {graphics} plotting functionality are also supported.
The focus is given to {ggplot2} as it is more verbose and efficient in creating data visualization based on "_The Grammar of Graphics_" [@wilkinson2012grammar].
The layered grammar makes developing charts more structural and easy to interpret (after you know how to use {ggplot2} of course).
One of the greatest strengths of {ggplot2} is the ability to customise it so easily.
Several themes and colour palettes already exist for {ggplot2} to Create the visualisation look professional and engaging for the end users.
{theme.aagi} leverages the ability to customise the appearance {ggplot2} to create a theme that is clean, easy to use and follows AAGI's style guidelines including fonts and colours.

When using {theme.aagi} for graphs, the legend will be placed at the top by default and the main and sub-titles will be left aligned and captions will be right aligned.
These choices can all be overridden by using `ggplot2::theme()` arguments as you wish.

## Getting started

### Install the {theme.aagi} package

{theme.aagi} is not on CRAN, so you will have to install it directly from our GitHub repository using {remotes}.

If you do not have the {remotes} package installed, you will have to run the first line in the code below as well. 

``` {r eval = FALSE}
if (!require("remotes"))
  remotes::install_github("CCDM_CBADA/theme.aagi",
                          build_vignettes = TRUE)
)
```

For more info on {theme.aagi} check out the [package's GitHub repository](https://github.com/CCDM_CBADA/theme.aagi), but most of the details about how to use the package and its functions are detailed below. 

When you have downloaded the package and successfully installed it you are good to go and create charts.

### Load the Libraries We'll Need

A few of the steps in this cookbook - and to create charts in R in general - require certain packages to be installed and loaded.
So that you do not have to load them one by one, you can use the the following code to load them all at once.

```{r libraries, eval=TRUE, message=FALSE}
x <-
  c("dplyr",
    "flextable",
    "gapminder",
    "ggplot2",
    "scales",
    "theme.aagi",
    "tibble")
invisible(lapply(
  X = x,
  FUN = library,
  character.only = TRUE
))
```

## Using {theme.aagi} to Create Tabular Outputs

{theme.aagi} provides a {flextable} them suited for the AAGI style.
You can use it like so.

```{r tabular}
ft <- flextable(head(airquality) |>
 mutate(`Month Name` = "May"))
ft <- theme_ft_aagi(ft)
ft
```

## Using {theme.aagi} to Create Graphical Outputs

{theme.aagi} provides several functions to assist users in creating plots, charts and graphs with a more unified AAGI style.
Following are examples of four major styles of graphs that are commonly used, bar plots, boxplots, histograms and lines and scatter plots.

### Bar Plots

#### Using Base R to Create a bar plot

```{r bar plot_base}
barplot_aagi(islands)
```

#### Using {ggplot2} to Create a Bar Plot

We need to transform a named vector to a `data.frame` for {ggplot2} to be able to use it, so we'll create a {tibble} first, `islands_df` and then plot it.

```{r bar plot_ggplot2}
islands_df <- tibble(name = names(islands), value = islands)

ggplot(data = islands_df, aes(x = name, y = value)) +
  geom_col() +
  theme_aagi() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Although those names are a bit difficult to read on the x-axis, so we can flip the coordinates so that they are easier to read.

```{r bar plot_ggplot2_flipped, fig.height = 6}
ggplot(data = islands_df, aes(x = name, y = value)) +
  geom_col() +
  theme_aagi() +
  coord_flip()
```

### Boxplots

#### Using Base R to Create a Boxplot

```{r boxplot-base}
boxplot_aagi(decrease ~ treatment,
             data = OrchardSprays,
             xlab = "treatment",
             ylab = "decrease")
```

#### Using {ggplot2} to Create a Boxplot

```{r boxplot-ggplot2}
library("theme.aagi")
library("ggplot2")

ggplot(data = OrchardSprays, aes(x = treatment, y = decrease)) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 120, by = 20)) +
  theme_aagi()
```

### Histograms

#### Using Base R to Create a Histogram

```{r hist-base}
library("ggplot2") # for diamonds data set
hist_aagi(diamonds$carat)
```

#### Using {ggplot2} to Create a Histogram

```{r hist-ggplot2}
ggplot(data = diamonds, aes(x = carat)) +
  geom_histogram() +
  theme_aagi()
```

### Line Graphs

Using the {gapminder} data, we can make the following figure of life expectancy over time in Malawi with the line in AAGI's turquoise colour.

#### Using Base R to Create a Line Graph

```{r line-base}
# Data for chart from {gapminder} package
line_df <- gapminder |>
  filter(country == "Malawi") |>
  arrange(year)

plot_aagi(
  x = line_df$year,
  y = line_df$lifeExp,
  col = aagi_cols("AAGI Turquoise"),
  type = "l",
  main = "Living longer",
  xlab = "Year",
  ylab = "Life expectancy",
  sub = "Life expectancy in Malawi 1952-2007"
)
```

#### Using {ggplot2} to Create a Line Graph

```{r line-ggplot2}
ggplot(line_df, aes(x = year, y = lifeExp)) +
  geom_line(colour = aagi_cols("AAGI Turquoise")) +
  theme_aagi() +
  ylab("Life expectancy") +
  xlab("Year") +
  labs(title = "Living longer",
       subtitle = "Life expectancy in Malawi 1952-2007")
```

### Scatterplots

```{r scatter-base-data}
# Create data
data = data.frame(
  x = seq(1:100) + 0.1 * seq(1:100) * sample(c(1:10), 100 , replace = TRUE),
  y = seq(1:100) + 0.2 * seq(1:100) * sample(c(1:10), 100 , replace = TRUE)
)
```

#### Using base R to Create a Scatterplot

```{r scatter-base}
# Basic scatterplot
plot_aagi(x = data$x, y = data$y)
```

#### Using {ggplot2} to Create a Scatterplot

```{r scatter-ggplot2}
ggplot(data = data, aes(x = x, y = y)) +
  geom_point() +
  theme_aagi()
```

## Colours in {theme.aagi}

### AAGI Colours

{theme.aagi} includes official AAGI colours for use in plots and also applies them to {ggplot2} facet strips and uses them in the MS Word and PowerPoint templates for colour matching in the outputs.

```{r colours}
library(scales)
show_col(aagi_colours)
```

### Using AAGI Colours

We've already seen above in the [line graph](#Linegraphs) example how to use the colours in a graph.
But for further demonstration, here are a few more examples.

Here we'll again use the {gapminder} data to construct a stacked bar chart and use AAGI's colours `scale_fill_aagi()`.

```{r aagi-facet-example}
#prepare data
stacked_df <- gapminder |>
  filter(year == 2007) |>
  mutate(lifeExpGrouped = cut(
    lifeExp,
    breaks = c(0, 50, 65, 80, 90),
    labels = c("Under 50", "50-65", "65-80", "80+")
  )) |>
  group_by(continent, lifeExpGrouped) |>
  summarise(continentPop = sum(as.numeric(pop)))

#set order of stacks by changing factor levels
stacked_df$lifeExpGrouped = factor(stacked_df$lifeExpGrouped,
                                   levels = rev(levels(stacked_df$lifeExpGrouped)))

#create plot
ggplot(data = stacked_df,
       aes(x = continent,
           y = continentPop,
           fill = lifeExpGrouped)) +
  geom_bar(stat = "identity",
           position = "fill") +
  scale_fill_aagi(name = "Life expectancy") +
  ylab("Continental Population (%)") +
  xlab("Continent") +
  theme_aagi() +
  scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = 0,
             linewidth = 1,
             colour = "#474747") +
  labs(title = "How life expectancy varies",
       subtitle = "% of population by life expectancy band, 2007") +
  guides(fill = guide_legend(reverse = TRUE))
```


## Advanced Uses

### Setting the AAGI Table Theme Globally

You can set your R session to use `theme_ft_aagi()` for {flextable} and {ggplot2} objects in your RMarkdown or Quarto documents to apply this theme to all tables using the following code chunk in your document at the top when setting up the options.

```{r eval=FALSE}
library(theme.aagi)

# set {ggplot2} theme
theme_set(theme_aagi())

# set {flextable} theme
set_flextable_defaults(
  theme_fun = theme_ft_aagi
)
```

### Using {ggplot2} Faceting

As we'll be using this figure for the next example, we'll generate an object in R called `p` and save it to R's `tempdir()` for use in the next example as well.

```{r ggplot2-facet, eval=TRUE}
#Prepare data
facet <- gapminder |>
  filter(continent != "Americas") |>
  group_by(continent, year) |>
  summarise(pop = sum(as.numeric(pop)))

col_values <- unname(c(
  aagi_cols("AAGI Turquoise"),
  aagi_cols("AAGI Green"),
  aagi_cols("AAGI Yellow"),
  aagi_cols("AAGI Blue")
))

#Make plot
p <- ggplot() +
  geom_area(data = facet, aes(x = year, y = pop, fill = continent)) +
  scale_fill_manual(values = col_values) +
  facet_wrap(~ continent, ncol = 5) +
  scale_y_continuous(breaks = c(0, 2000000000, 4000000000),
                     labels = c(0, "2bn", "4bn")) +
  theme_aagi() +
  geom_hline(yintercept = 0,
             linewidth = 1,
             colour = "#474747") +
  theme(legend.position = "none", axis.text.x = element_blank()) +
  labs(title = "Asia's rapid growth", subtitle = "Population growth by continent, 1952-2007")

ggsave(p, filename = "AAGI.png", path = tempdir())

print(p)
```

### Adding the AAGI Logo to Your Figures

Add the AAGI logo to the upper left of the previous plot as per the style guide.
Use the `add_aagi_logo()` function to add the logo automatically to a previously saved image file.
In this case, the previous example used to show the faceting colours has been saved to R's `tempdir()`.
You can save using `tempdir()` as illustrated above or save on your Documents folder or elsewhere and access it from there as well.
The image is resized to be 600x1000 pixels before being displayed here.
Feel free to adjust as necessary.

```{r add-dpird-logo, eval=TRUE, message=FALSE}
library(magick)
add_aagi_logo(file_in = file.path(tempdir(), "AAGI.png"),
               file_out = file.path(tempdir(), "AAGI_logo.png"))

image_read(file.path(tempdir(), "AAGI_logo.png")) |> 
  image_resize("600x1000") |> 
  print()
```

## Saving {ggplot2} Figures

When saving figures generated with {ggplot2} that use {theme.aagi} as the theme, PDFs will not properly embed the Proxima Nova font that is used.
In order for you to save your figures and preserve the font, you can either save as a .png file or use the `device` option in `ggsave()`, which is demonstrated as follows.

```{r save-ggplot, eval=TRUE}
ggsave(
  filename = "cairo_font.pdf",
  plot = p,
  device = cairo_pdf,
  path = tempdir()
)
```

### Cairo Instructions for macOS

You should already have the Cairo graphics library installed when you install R, unless you use the Homebrew version, _e.g._, `brew install r`.
You can verify that you have Cairo support by running the `capabilities()` function; `TRUE` should show up under cairo:

```{r r-capabilities}
capabilities()
```

If you do not see `TRUE` under "cairo", check that you've installed a version of R using the Homebrew cask, different than `brew install r`, this installs the vanilla version of R as built by the R team or using rig, <https://github.com/r-lib/rig>, to manage your R installation if you're not just installing the default R binary.

So, you should not need to install any R-specific Cairo libraries or anything for this to work.
However, you will need an X11 window system first, like XQuartz.
You can install XQuartz like so.

```bash
$ brew install xquartz
```

### Cairo Instructions for Windows

You will already have the Cairo graphics library installed when you install R.
There is no need to do anything else to Create it work on this platform.
